---
title: Predicting Euro & Copa America - 2024
description: A statistical model to analyze the results of the Euro Cup and the Copa America 2024.
author: 
  - name: Mauro Florez
  -  email: mf53@rice.edu
date: 2024-06-30
bibliography: references.bib
---

In this project, I will propose a model for predicting the Euro & Copa America - 2024 scores based on the model presented in [@florez2024].

## Packages

I will use a Bayesian model, so we need the following libraries

```{r, message=FALSE}
require(rstanarm)
require(rstantools)
require(bayesplot)
```

As well as

```{r, message=FALSE}
require(readr)
require(dplyr)
require(lubridate)
require(ggplot2)
require(ggrepel)
require(latex2exp)
```

## Data

The data is contained in this repository <https://github.com/martj42/international_results>.

```{r output = FALSE}
urlfile = "https://raw.githubusercontent.com/martj42/international_results/master/results.csv"
```

We load the data

```{r }
data <- read_csv(url(urlfile), show_col_types = FALSE)
```

The proposed model is based on the teams' strengths attacking and defending. To train the model, we used the scores of games played since season 2023-2024.

```{r}
euro_2024 = c('Albania','Austria','Belgium','Croatia','Czech Republic','Denmark','England','France','Georgia','Germany','Hungary','Italy','Netherlands','Poland','Portugal','Romania','Scotland','Serbia','Slovakia','Slovenia','Spain','Switzerland','Turkey','Ukraine')

america_2024 = sort(c('Argentina', 'Chile', 'Peru', 'Canada', 'Mexico', 'Ecuador', 'Venezuela', 'Jamaica', 'United States', 'Uruguay', 'Panama', 'Bolivia', 'Brazil', 'Colombia', 'Paraguay', 'Costa Rica'))  

scores <- data %>% 
  filter(home_team %in% euro_2024 & away_team %in% euro_2024) %>% 
  filter(date > "2023-06-01") %>% 
  mutate(id = row_number(), cup = "Euro") %>% 
  bind_rows(data %>% 
  filter(home_team %in% america_2024 & away_team %in% america_2024) %>% 
  filter(date > "2023-06-01") %>% 
  mutate(id = row_number(), cup = "America")
  )
```

For example, we consider the following games played by Portugal.

```{r}
scores %>% 
  filter(home_team == "Portugal" | away_team == "Portugal")
```

and the following games played by Colombia

```{r}
scores %>% 
  filter(home_team == "Colombia" | away_team == "Colombia")
```

### Filtering data

```{r}
home_scores <- scores %>% 
  select(id, score = home_score, team = home_team, adversary = away_team, date = date, cup) %>% 
  mutate(condition = ifelse(scores$neutral, "neutral", "home"))
  
away_scores <- scores %>% 
  select(id, score = away_score, team = away_team, adversary = home_team, date = date, cup) %>% 
  mutate(condition = ifelse(scores$neutral, "neutral", "away"))

total_scores <- union_all(home_scores, away_scores)
```

We also consider the ranking FIFA of the teams as covariate. For every game, select the up-to-date ranking

```{r}
path0 <- "https://raw.githubusercontent.com/Dato-Futbol/fifa-ranking/master/ranking_fifa_historical.csv"
rank <- read_csv(url(path0), show_col_types = FALSE)

rank <- rank %>% 
  select(team, total_points, date) %>% 
  group_by(date) %>% 
  mutate(ranking = rank(-total_points))
```

Merging the data will be

```{r}
scores_rank <- total_scores %>% 
  left_join(rank, by = join_by(adversary == team), relationship = "many-to-many") %>% 
  filter(date.x >= date.y) %>% 
  mutate(diff_date = date.x - date.y) %>% 
  group_by(id, team) %>% 
  slice_min(diff_date) %>% 
  select(id, score, team, adversary, condition, date_game = date.x, ranking_adv = ranking, date_rank = date.y, cup)
```

We are separating the data for both Euro and Copa America. For both cases, we will use as test data the games played after the quarter-finals. i.e, we will finish predicting the results of the semi-finals and the finals.

```{r}
date_euro = "2024-07-09"
date_ca = "2024-07-09"

train_euro <- scores_rank %>% 
  filter(date_game < date_euro, cup == "Euro")

train_america <- scores_rank %>% 
  filter(date_game < date_ca, cup == "America")

test_euro <- scores_rank %>% 
  filter(date_game >= date_euro, cup == "Euro")

test_america <- scores_rank %>% 
  filter(date_game >= date_ca, cup == "America")
```

## Fitting the model

To fit the model I would consider the scores are distributed as Poisson for simplicity, one better option is the Conway-Maxwell-Poisson as shown in [@florez2024]. However, we will use a simple model for computational reasons as we constantly update the model.

```{r, output=FALSE}
model_euro <- stan_glm(score ~ 1 + team + adversary + condition + ranking_adv, data = train_euro, family = poisson, iter = 5000, contrasts = list(team = "contr.sum", adversary = "contr.sum"))
```

```{r, output=FALSE}
model_america <- stan_glm(score ~ 1 + team + adversary + condition + ranking_adv, data = train_america, family = poisson, iter = 5000, contrasts = list(team = "contr.sum", adversary = "contr.sum"))
```

## Estimation

### Euro

```{r}
strength_att <- model_euro$coefficients[2:24]
strength_att <- c(strength_att, 0 - sum(strength_att))

strength_def <- model_euro$coefficients[25:47]
strength_def <- c(strength_def, 0 - sum(strength_def))

coeffs <- data.frame(team = euro_2024, strength_att, strength_def, row.names = c())
```

Now, we can plot the estimated strengths in the following way:

```{r}
ggplot(coeffs, aes(x = strength_att, y = -strength_def)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point() +
  geom_label_repel(aes(label = team),
                            box.padding = 0.25,
                            point.padding = 0.5,
                            segment.color = "grey50") +
  xlab(expression("Better Attack" %->% "")) +
  ylab(expression("Better Defense" %->% "")) + 
  ggtitle("Strengths of every team") + 
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```

This graph provides an intuition into the strengths and abilities of every team. Teams located in the origin have an overall performance, and below the axis are worse than the averages' team.

### Copa America

Similarly for Copa america

```{r}
n <- length(america_2024)
strength_att <- model_america$coefficients[2:n]
strength_att <- c(strength_att, 0 - sum(strength_att))

strength_def <- model_america$coefficients[(n+1):(2*n-1)]
strength_def <- c(strength_def, 0 - sum(strength_def))

coeffs <- data.frame(team = america_2024, strength_att, strength_def, row.names = c())
```

Now, we can plot the estimated strengths in the following way:

```{r}
ggplot(coeffs, aes(x = strength_att, y = -strength_def)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point() +
  geom_label_repel(aes(label = team),
                            box.padding = 0.25,
                            point.padding = 0.5,
                            segment.color = "grey50") +
  xlab(expression("Better Attack" %->% "")) +
  ylab(expression("Better Defense" %->% "")) + 
  ggtitle("Strengths of every team") + 
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```

## Predictions

### Euro

Now, we predict the scores of the future games, i.e, the semifinals and the final.

```{r}
test_pred <- test_euro %>% 
  select(id, team, adversary, condition, ranking_adv)

predictions <- posterior_predict(model_euro, newdata = test_pred) 

mean_goals <- apply(predictions, 2, mean)

test_pred <- test_pred %>% 
  ungroup() %>% 
  mutate(goals = mean_goals) %>% 
  group_by(id) %>% 
  summarise("Team 1" = first(team),
            "Exp. Goals" = first(goals),
            "Team 2" = last(team),
            "Exp. Goals 2" = last(goals)) 
```

The predictions of future games are:

```{r}
test_pred %>% 
  print(n = 1e3)
  
```

Particullarly, the distribution of the prediction of the score for the final looks as shown below.

```{r}
results_game <- tibble("England" = predictions[,5], "Spain" = predictions[,6])

df_count <- results_game %>% 
  group_by(England, Spain) %>% 
  summarise(count = n(), .groups = "keep") %>% 
  ungroup()

total_combinations <- nrow(results_game)

df_count <- df_count %>% 
    mutate(proportion = count/total_combinations)
  
ggplot(df_count, aes(x = England, y = Spain, fill = proportion)) +
  geom_tile() + 
  scale_fill_gradient(low = "white", high = "blue") + 
  labs(title = "Heatmap of Goals Scored by the two teams",
       x = "England", 
       y = "Spain",
       fill = "Probability") + 
  theme_minimal()
```

According to this, it is likely that the final will be Spain vs. England. **Spain is expected to be the champion, with a predicted score of 2-0.**

### Copa America

Similarly, we can generate the predictions for the last round of the games in Copa América

```{r}
test_pred <- test_america %>% 
  select(id, team, adversary, condition, ranking_adv)

predictions <- posterior_predict(model_america, newdata = test_pred) 

mean_goals <- apply(predictions, 2, mean)

test_pred <- test_pred %>% 
  ungroup() %>% 
  mutate(goals = mean_goals) %>% 
  group_by(id) %>% 
  summarise("Team 1" = first(team),
            "Exp. Goals" = first(goals),
            "Team 2" = last(team),
            "Exp. Goals 2" = last(goals)) 

```

```{r}
test_pred %>% 
  print(n = 1e3)
```

In this case, Argentina is expected to easily defeat Canada, and Uruguay will likely beat Colombia. However, Colombia won its game 1-0, so the final will be Argentina vs. Colombia, while the 3rd and 4th place game will be played between Canada and Uruguay.

**The final is anticipated to be a close match, with the most likely result being 0-0, as indicated below. However, Argentina is expected to have a slightly higher probability of scoring more goals than Colombia.**

```{r}
results_game <- tibble("Argentina" = predictions[,7], "Colombia" = predictions[,8])

df_count <- results_game %>% 
  group_by(Argentina, Colombia) %>% 
  summarise(count = n(), .groups = "keep") %>% 
  ungroup()

total_combinations <- nrow(results_game)

df_count <- df_count %>% 
    mutate(proportion = count/total_combinations)
  
ggplot(df_count, aes(x = Argentina, y = Colombia, fill = proportion)) +
  geom_tile() + 
  scale_fill_gradient(low = "white", high = "blue") + 
  labs(title = "Heatmap of Goals Scored by the two teams",
       x = "Argentina", 
       y = "Colombia",
       fill = "Probability") + 
  theme_minimal()
```

As we saw, the model accurately predicted the scores and captured the strengths of each team. It also produced reasonable and accurate results for the final games based on the quarter-final outcomes. **Spain defeated England 2-1**, while the match between **Argentina and Colombia ended 0-0**, with Argentina scoring a goal in extra time.

For future models, incorporating dynamic strengths could improve accuracy.
